SPENCER THESIS - WORK IN PROGRESS

### 2016-06-04 doc/thesis_swe.tex
Thesis, figure demo, and test update. #98

\begin{enumerate}
\item Under the best possible circumstances of having noiseless redshifts and halo masses, how accurately can I reconstruct line-of-sight dark matter mass structures using a halo model?
%\item What fraction of dark matter mass is in halos?
\item How many halos are needed to make these predictions?
\item How feasible is an inference of the Millennium Simulation's cosmological hyper-parameters, such as the stellar mass - halo mass relation, using all available halo data? 
\item How long does such an inference take?
\end{enumerate}
%Guided by these results, in this thesis I build upon the work of Collet et al. in \cite{collet} to apply a simple dark matter halo model to reconstruct lensing mass along lines of sight in the Millennium Simulation using the \texttt{Pangloss} framework in order to predict the weakly lensed ellipticities of generated background sources.

@@

Cosmological simulations like the Millennium Simulation (MS) demonstrate that the underlying matter distribution of our universe is likely breathtakingly complex; rather than a largely disordered collection of galaxies and interstellar gas, matter appears to condense into clusters and superclusters of galaxies threaded together by large filaments pocketed with under-dense regions called voids, as shown in Figure \ref{fig:ms_structure} (Cite MS). While the exact relation between the distribution of galaxies and dark matter is not known, simulations predict that early galaxies formed in the large gravitational potentials of over-dense regions of dark matter (Citation). This suggests that, while we cannot map dark matter directly, galaxies should trace out at least some of the underlying dark matter structures (Citation). A comparrison of the predicted galaxy and dark matter distribution from a small region of the MS is shown in Figure \ref{ms_compare}.

@@

The simplest way to model the relationship between galaxies and dark matter is by enveloping each galaxy in a spherically symmetric dark matter `halo' of a certain mass $M_h$. These halos extend far beyond the edge of the visible galaxy that they enclose, with the Milky Way's own halo estimated to have a radius of (Check and cite). While the density profile of the halos may be complex, numerous simulations have shown that it can be well approximated locally by the Navarro-Frenk-White (NFW) profile which has the form
\begin{equation}\label{nfw_profile}
\rho_{NFW}(r)=\frac{\rho_0}{\frac{r}{R_s}\left(1+\frac{r}{R_s}\right)^2}
\end{equation}

@@

\begin{align}\label{rho_h}
\rho_h(z)&=\frac{\sum_i M_{h,i}}{V(z)}\nonumber\\
&=\sum_i M_{h,i}\cdot\left(d_H\int\frac{d_M(z^\prime)^2}{E(z^\prime)(1+z^\prime)^3}dz^\prime d\Omega\right)^{-1}.
\end{align}

@@

While dark matter does not absorb or emit light, it does interact with light indirectly through gravitational lensing. The presence of matter curves spacetime locally which causes light to follow curved paths called geodesics. The deviation in path is usually negligable, but can be significant in regions containing clusters of galaxies and their massive dark matter halos. By observing how the magnitudes and shapes of distant galaxies are distorted when travelling through regions of matter, we can infer the masses of the dark matter halos and reconstruct the hidden dark matter mass distribution. A simplified cartoon of this process is shown in Figure ??, as well as a diagram defining the necessary angles and distances in Figure ??.

@@

\noindent where $\sum(\vec{r},z_l)$ is the projected surface mass density of the lensing mass and $\sum_{cr}(z_l,z_s)$ is the critical surface density for a given source redshift $z_s$ and lens redshift $z_l$ (Citation). The shearing of the source is most often described by the complex shear $\gamma$ defined to be
\begin{equation}\label{complex_shear}
\gamma=\gamma_1+i\gamma_2=|\gamma|\,e^{2i\varphi}
\end{equation}

@@

 The analytical form of the convergence and shear contribution from a BMO-profile is quite messy, but is derived and explained in detail in \ref{bmo}. Calculating the lensing contribution of the smooth-component correction is much simpler; the projected surface mass density at at given redshift is found by integrating $\rho_s(z)$ over a sufficiently small $dz$ interval, and then the mass sheet's convergence is found using Equation \eqref{kappa}. There is no shear contribution as a uniform sheet of mass causes only magnification due to the mass sheet degeneracy (CITE!). However, Equation \eqref{lensing} shows that the lensed ellipticities depend on the \textit{reduced} shear, which does depend on the additional convergence terms resulting from the smooth-component correction.

@@

If all galaxies were spherical then the measurement of their ellipticities would give an exact description of the lensing done by foreground mass structures. Unfortunately, the intrinsic ellipticities of background galaxies contribute significant noise as we do not know how any individual galaxy ellipticity has been altered by lensing. However as two sources lensed by the same foreground mass will have their ellipticities distorted in a similar way, their lensed ellipticities should be correlated regardless of intrinsic orientation. The average correlation of a sufficient number of galaxy pairs should overcome the noise and lead to a detectable signal. Importantly, this implies that \textit{the correlation of galaxy ellipticities is a probe of the underlying dark matter mass distribution} (Citation).

The mathematical tool that measures the average correlation as a function of separation distance between two galaxies is called the ellipticity-ellipticity ($\varepsilon$-$\varepsilon$) correlation function. A quick overview of this function is given below, but readers that are unfamiliar with correlation functions in the context of cosmology or want a visual aid should see Appendix (??).

@@

Not all foreground halos are equally relevant to the lensing calculation. To add a significant contribution to the combined shear and convergence, a halo must be massive, close to the line of sight, or preferably both. McCully et al. (Cite) derived that the correct metric for determining the relevance of a given foreground halo to the overall lensing calculation to be given by
\begin{equation}\label{relevant}
Rel(M_h,R)\propto\frac{M_h}{R^3}
\end{equation}

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
%% Smooth-Component Correction Diagram
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{figs-swe/thesis/smooth_correction.png}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{CAPTION}
    \label{fig:scc}
\end{figure}
%%------------------------------------------------------------------------------------------------------------------------------------------

@@

\textcolor{red}{Talk about lensing table!!! ??}

@@

Instead of comparing the ray-traced and \texttt{Pangloss} predicted ellipticities for individual galaxies, the lensing is characterized globally with correlation functions. As described in section \ref{corr_functions}, the ellipticity-ellipticity correlation function measures how correlated the ellipticities of pairs of galaxies are as a function of separation distance, while the galaxy-mass correlation function measures the correlation of lensed ellipticities around foreground halos as a function of separation distance. Both correlation functions are used in this work to estimate how well the \texttt{Pangloss} framework models weak lensing by dark matter structures and are computed using the publicly available \texttt{TreeCorr} module written by Mike Jarvis\footnote{https://github.com/rmjarvis/TreeCorr}. Note that the ellipticity-ellipticity correlation function definition used in \texttt{TreeCorr} is slightly different than that used in most of the literature; for a derivation of the connection between Jarvis's definition and the more common Schneider definition \cite{schneider}, see Appendix (??).

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
% Ellipticity-Ellipticity Correlation
\begin{figure}
    \centering
    \begin{subfigure}{0.475\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/gg_corr.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{CAPTION}
        %The ellipticity-ellipticity correlation function for ?? sources at separation distances from 0.1 to 2 arcminutes and a lightcone radius of $R=2$ arcminutes. While the correlation of the intrinsic ellipticities before lensing (blue) shows no correlation, the correlation of the ray-traced ellipticities (green) and halo model predicted ellipticities (purple) both show significant positive correlation on scales up to 1 arcminute. The halo model predicts a small negative correlation above this scale, while the ray-tracing shows positive correlation on all scales.}
        %The correlation of the intrinsic ellipticities before lensing is plotted in blue, and shows no correlation. The correlation of the true ray-traced ellipticities is plotted in green and shows significant positive correlation on all scales. The halo predicted correlation, plotted in purple, shows significant positive correlation until a separation distance of 0.8 arcminutes.
        \label{gg_corr}
    \end{subfigure}
    ~~
    \begin{subfigure}{0.475\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/gg_progression.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{CAPTION}
        %The ellipticity-ellipticity correlation functions for a series of halo model frameworks with increasing lightcone radii from $R=1$ to $R=8$ arcminutes, all plotted in purple with various line styles. These are compared to the same ray-traced correlation from Figure \ref{gg_corr}.}
        \label{gg_corr_series}
    \end{subfigure}
    \caption{}
\end{figure}
%------------------------------------------------------------------------------------------------------------------------------------------

The first test of the halo model was with the $\varepsilon$-$\varepsilon$ correlation function with a lightcone radius of 6 arcminutes which is given in Figure \ref{gg_corr}. The statistic measured the average correlation between pairs of ellipticities at separation distances between 0.1 and 2 arcminutes. The cross-component of the correlation function shows no significant deviation from zero as expected, as gravitational lensing only shears galaxies tangentially. For the $\xi_+$ component, the ray-traced (plotted in green) and halo model (plotted in purple) are in relative agreement from seperation scales of 0.1 to 0.2 arcminutes, but there is a clear systematic underprediction of correlation by the halo model on scales larger than 0.2 arcminutes. This may be indicative that not enough halo mass is being used in the lensing prediction, the halo model does not adequately address large-scale mass structures, or that there is significant dark matter mass not correlated with galaxies. These issues are discussed further in Section \ref{issues}.

Using the same catalog of background sources, this statistic was recalculated for the halo model ellipticities for various lightcone radii ranging from $R_L=1$ to $R_L=8$ arcminutes. The result is shown in Figure \ref{gg_corr_series}, where the series of model predicted correlation functions is compared to the ray-traced correlation function. As $R_L$ increases, the mean number of foreground galaxies contained in each lightcone grows quadratically from ${134\pm17}$ galaxies when $R_L=1$ arcminute to ${8570\pm87}$ galaxies when $R_L=8$ arcminutes. Increasing the number of foreground objects, and thus increasing the mass and structure considered for shear and convergence predictions, systematically increased the predicted correlation on all scales. This result encourages the use of lightcones with radii of at least 8 arcminutes, and further emphasizes the need for computational efficiency to allow the timely use of large lightcones.

@@

The catalog of lensed ellipticities was also analyzed using the galaxy-mass correlation function with separation distances from 0.1 to 2 arcminutes and a lightcone radius of 6 arcminutes. The result is shown in Figure \ref{ng_corr}. The lensed ellipticities from the ray-tracing and halo model both show positive correlation on all calculated scales and largely agree on scales larger than 0.3 arcminutes except for a slight overprediction by the halo model near 0.45 arcminutes. However, a new feature is the large overprediction of correlation on separation scales smaller than 0.2 arcminutes. This is likely because background sources this close to massive foreground halos are often in the strong lensing regime which is not currently accounted for by the \texttt{Pangloss} framework.

The galaxy-mass correlation function was similarly re-calculated at increasing lightcone radii from $R=1$ to $R=8$ arcminutes with the result in Figure \ref{ng_corr_series} As with the $\varepsilon$-$\varepsilon$ correlation, structure larger than $\sim$1 arcminute is captured much better by increasingly large lightcone radii as expected. However, there is little to no improvement of the galaxy-mass correlation on scales smaller than 0.3 arcminutes which futher points to the need to model strong lensing.

%%------------------------------------------------------------------------------------------------------------------------------------------
%% Galaxy-mass correlation function
\begin{figure}
    \centering
    \begin{subfigure}{0.475\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/ng_corr.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{CAPTION}
        %The galaxy-mass correlation function of 3,240 sources at separation distances from 0.1 to 2 arcminutes and a lightcone radius of $R=2$ arcminutes. There is no significant correlation of the intrinsic ellipticities (blue) as expected, and both the ray-traced correlation (green) and halo model correlation (purple) are positive on all scales. The green and purple curves are in agreement from 0.2 to 0.7 arcminutes, but have significant discrepancies on scales above and below this range.}
        \label{ng_corr}
    \end{subfigure}
    ~~
    \begin{subfigure}{0.475\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/ng_progression.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{CAPTION}
        \label{ng_corr_series}
    \end{subfigure}
    \caption{}
\end{figure}
%
%%------------------------------------------------------------------------------------------------------------------------------------------

@@

Of intrinsic interest is the relative number of relevant halos needed in each lightcone to accurately predict the halo correlation functions in Figures \ref{fig:gg_corr} and \ref{fig:ng_corr}, as a low percentage would allow much faster lensing predictions. Figure \ref{fig:rel_corr} displays an example $\varepsilon$-$\varepsilon$ and galaxy-mass correlation function comparrison of the halo model using all halos (purple) and halo model using halos above the relevance limit of ${10^{-5} M_{\odot}/\text{Mpc}^3}$, as well as the percent error of each predicted value between the two implementations. This relevance limit resulted in $??\pm??$ relevant halos per lightcone, and resulted in a normalized root-mean-square error (NRMSE) of ??. Figure \ref{fig:halo_fraction} plots the $\varepsilon$-$\varepsilon$ and galaxy-mass correlation function NRMSE as a function of relevance limits and mean relevant halos. These results suggest that either correlation function prediction can be approximated with an NRMSE below 10\% using around the ?? most relevant halos, or an NRMSE of below 5\% using around the ?? most relevant halos. Calibration curves such as this can allow the best-case correlation function to be predicted within a desired accuracy at only a fraction of the full lensing calculation which will be especially helpful in future analyses with much larger fields of view.

@@

%------------------------------------------------------------------------------------------------------------------------------------------
% NRMSE and Correlation Fraction as function of Mean Relevant Halos
\begin{figure}
    \centering
    \includegraphics[width=0.475\textwidth]{figs-swe/thesis/halo_fraction.png}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{CAPTION!}
    \label{fig:halo_fraction}
\end{figure}

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Dark Matter Convergence Mass Maps}

@@

The result of a particular ?? arcmin$^2$ field is shown in Figure \ref{fig:kappamaps}, where the ray-traced convergence is shown on the left, the halo model convergence is in the center, and the ray-traced convergence subtracted from the halo model convergence is shown on the right. While particular foreground masses are almost always predicted, the currently implemented halo model over-predicts the convergence throughout most of the field. The halo model does not appear to accurately account for the voids between foreground masses and over-predicts the mass of most ... This could partially be a result of incorrect halo profiles, but it is apparent that the smooth-component correction is not adequate to ...

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Computational Performance}

A secondary goal of this research was to investigate how quickly the prediction for the shear and convergence for a single background source could be made. The mean CPU time required for the prediction per lightcone was ${??\pm??}$ ms for the first run. However, many small improvements to the coding infrastructure of \texttt{Pangloss} were made to speed up the prediction. The average CPU time for a lensing prediction per lightcone for ?? separate runs after various (and cumulative) changes is plotted in Figure \ref{cpu_plot}, with a final run time of $??\pm?$ ms. Run 1 corresponds to the original \texttt{Pangloss} framework code, run 2 only wrote the needed quantities for this analysis to the catalog \texttt{Astropy} tables, run 3 changed the redshift grid binning to be done once for all halos in the drilling rather than for every individual halo in lensing, run 4 optimized various lensing calculations with matrices whenever possible, run 5 replaced the computationally expensive lensing calculation functions with a lookup table, and run 6 is using only the most ?? relevant halos. Cumulatively, these changes alone accounted for a ??\% decrease in mean CPU time per lightcone and ?? the initial variation.

The distribution of total CPU time in significant parts of the lensing calculation for run 6 are as follows: 24.0\% of total time calculating individual halo convergence contributions where 14.4\% is spent in lensing lookup tables, 65.1\% of CPU time spent combining all convergence contributions where 62.1\% is spent on the smooth-component correction, and the majority of the remaining time spent on writing \texttt{Astropy}\footnote{http://www.astropy.org/} and \texttt{NumPy}\footnote{http://www.numpy.org/} columns.

While progress has been made, the code needs to be further optimized to allow lensing predictions of fields of many square degrees to be done in a few seconds. Section \ref{scaling_up} discusses the details of future computational scale-up plans.

%%------------------------------------------------------------------------------------------------------------------------------------------
% Mean lightcone CPU time
\begin{figure}
    \centering
    \includegraphics[width=0.475\textwidth]{figs-swe/thesis/profile_bar.png}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{OLD: The mean CPU runtime per lightcone for various runs with different (and cumulative) code optimizations described in the text. The cumulative changes account for a ??\% decrease in the mean CPU time and a ?? of the variation. Currently, the prediction can be made for a single lightcone in\\$??\pm??$ ms.}
    \label{fig:cpu_plot}
\end{figure}
%%------------------------------------------------------------------------------------------------------------------------------------------

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Results}

>>>>

\begin{enumerate}
\item Under the best possible circumstances of having noiseless redshifts and halo masses, how accurately can I reconstruct line-of-sight dark matter mass structures using a halo model?
%\item What fraction of dark matter mass is in halos?
\item How many halos are needed to make these predictions?
\item How feasible is an inference of the Millennium Simulation's cosmological hyper-parameters, such as the stellar mass - halo mass relation, using all available halo data? 
\item How long does such an inference take?
\end{enumerate}

@@

Cosmological simulations like the Millennium Simulation (MS) demonstrate that the underlying matter distribution of our universe is likely breathtakingly complex; rather than a largely disordered collection of galaxies and interstellar gas, matter appears to condense into clusters and superclusters of galaxies threaded together by large filaments pocketed with under-dense regions called voids, as shown in Figure \ref{fig:ms_structure} \cite{??}. While the exact relation between the distribution of galaxies and dark matter is not known, simulations predict that early galaxies formed in the large gravitational potentials of over-dense regions of dark matter \cite{??}. This suggests that, while we cannot map dark matter directly, galaxies should trace out at least some of the underlying dark matter structures \cite{??}. A comparrison of the predicted galaxy and dark matter distribution from a small region of the MS is shown in Figure \ref{ms_compare}.

@@

The simplest way to model the relationship between galaxies and dark matter is by enveloping each galaxy in a spherically symmetric dark matter `halo' of a certain mass $M_h$. These halos extend far beyond the edge of the visible galaxy that they enclose, with the Milky Way's own halo estimated to have a radius of ?? \cite{??}. While the density profile of the halos may be complex, numerous simulations have shown that it can be well approximated locally by the Navarro-Frenk-White (NFW) profile which has the form
\begin{equation}\label{nfw_profile}
\rho_{NFW}(r)=\frac{\rho_0}{\frac{r}{R_s}\left(1+\frac{r}{R_s}\right)^2}
\end{equation}

@@

\begin{align}\label{rho_h}
\rho_h(z)&=\frac{\sum_i M_{h,i}}{V(z)}\nonumber\\
&=\sum_i M_{h,i}\cdot\left(d_H\int\frac{d_M(z^\prime)^2}{E(z^\prime)(1+z^\prime)^3}dz^\prime d\Omega\right)^{-1}
\end{align}

@@

While dark matter does not absorb or emit light, it does interact with light indirectly through gravitational lensing. The presence of matter curves spacetime locally which causes light to follow curved paths called geodesics. The deviation in path is usually negligable, but can be significant in regions containing clusters of galaxies and their massive dark matter halos. By observing how the magnitudes and shapes of distant galaxies are distorted when travelling through regions of matter, we can infer the masses of the dark matter halos and reconstruct the hidden dark matter mass distribution. A simplified cartoon of this process is shown in Figure ??.

@@

\noindent where $\sum(\vec{r},z_l)$ is the projected surface mass density of the lensing mass and $\sum_{cr}(z_l,z_s)$ is the critical surface density for a given source redshift $z_s$ and lens redshift $z_l$ \cite{??}. The shearing of the source is most often described by the complex shear $\gamma$ defined to be
\begin{equation}\label{complex_shear}
\gamma=\gamma_1+i\gamma_2=|\gamma|\,e^{2i\varphi}
\end{equation}

@@

 The analytical form of the convergence and shear contribution from a BMO-profile is quite messy, but is derived and explained in detail in \ref{bmo}. Calculating the lensing contribution of the smooth-component correction is much simpler; the projected surface mass density at at given redshift is found by integrating $\rho_s(z)$ over a sufficiently small $dz$ interval, and then the mass sheet's convergence is found using Equation \eqref{kappa}. There is no shear contribution as a uniform sheet of mass causes only magnification due to the mass sheet degeneracy \cite{??}. However, Equation \eqref{lensing} shows that the lensed ellipticities depend on the \textit{reduced} shear, which does depend on the additional convergence terms resulting from the smooth-component correction.

@@

If all galaxies were spherical then the measurement of their ellipticities would give an exact description of the lensing done by foreground mass structures. Unfortunately, the intrinsic ellipticities of background galaxies contribute significant noise as we do not know how any individual galaxy ellipticity has been altered by lensing. However as two sources lensed by the same foreground mass will have their ellipticities distorted in a similar way, their lensed ellipticities should be correlated regardless of intrinsic orientation. The average correlation of a sufficient number of galaxy pairs should overcome the noise and lead to a detectable signal. Importantly, this implies that \textit{the correlation of galaxy ellipticities is a probe of the underlying dark matter mass distribution} \cite{??}.

The mathematical tool that measures the average correlation as a function of separation distance between two galaxies is called the ellipticity-ellipticity ($\varepsilon$-$\varepsilon$) correlation function. A quick overview of this function is given below, but readers that are unfamiliar with correlation functions in the context of cosmology or want a visual aid should see Appendix \ref{visualize_corr}.

@@

Not all foreground halos are equally relevant to the lensing calculation. To add a significant contribution to the combined shear and convergence, a halo must be massive, close to the line of sight, or preferably both. McCully et al. in \ref{mccully} derived that the correct metric for determining the relevance of a given foreground halo to the overall lensing calculation to be given by
\begin{equation}\label{relevant}
Rel(M_h,R)\propto\frac{M_h}{R^3}
\end{equation}

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
%% Smooth-Component Correction Diagram
\begin{figure}
    \centering
    \includegraphics[width=0.475\textwidth]{figs-swe/thesis/smooth_correction.png}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{CAPTION}
    \label{fig:scc}
\end{figure}
%%------------------------------------------------------------------------------------------------------------------------------------------

@@

@@

Instead of comparing the ray-traced and \texttt{Pangloss} predicted ellipticities for individual galaxies, the lensing is characterized globally with correlation functions. As described in section \ref{corr_functions}, the ellipticity-ellipticity correlation function measures how correlated the ellipticities of pairs of galaxies are as a function of separation distance, while the galaxy-mass correlation function measures the correlation of lensed ellipticities around foreground halos as a function of separation distance. Both correlation functions are used in this work to estimate how well the \texttt{Pangloss} framework models weak lensing by dark matter structures and are computed using the publicly available \texttt{TreeCorr} module written by Mike Jarvis\footnote{https://github.com/rmjarvis/TreeCorr}. Note that the ellipticity-ellipticity correlation function definition used in \texttt{TreeCorr} is slightly different than that used in most of the literature; for a derivation of the connection between Jarvis's definition and the more common Schneider definition \cite{schneider}, see Appendix \ref{corr_defs}.

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
% Ellipticity-Ellipticity Correlation
\begin{figure}
    \centering
    \begin{subfigure}{0.475\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/gg_corr.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{CAPTION}
        %The ellipticity-ellipticity correlation function for ?? sources at separation distances from 0.1 to 2 arcminutes and a lightcone radius of $R=2$ arcminutes. While the correlation of the intrinsic ellipticities before lensing (blue) shows no correlation, the correlation of the ray-traced ellipticities (green) and halo model predicted ellipticities (purple) both show significant positive correlation on scales up to 1 arcminute. The halo model predicts a small negative correlation above this scale, while the ray-tracing shows positive correlation on all scales.}
        %The correlation of the intrinsic ellipticities before lensing is plotted in blue, and shows no correlation. The correlation of the true ray-traced ellipticities is plotted in green and shows significant positive correlation on all scales. The halo predicted correlation, plotted in purple, shows significant positive correlation until a separation distance of 0.8 arcminutes.
        \label{fig:gg_corr}
    \end{subfigure}
    ~~
    \begin{subfigure}{0.475\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/gg_progression.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{CAPTION}
        %The ellipticity-ellipticity correlation functions for a series of halo model frameworks with increasing lightcone radii from $R=1$ to $R=8$ arcminutes, all plotted in purple with various line styles. These are compared to the same ray-traced correlation from Figure \ref{gg_corr}.}
        \label{fig:gg_corr_series}
    \end{subfigure}
    \caption{}
\end{figure}
%------------------------------------------------------------------------------------------------------------------------------------------

The first test of the halo model was with the $\varepsilon$-$\varepsilon$ correlation function with a lightcone radius of 6 arcminutes which is given in Figure \ref{fig:gg_corr}. The statistic measured the average correlation between pairs of ellipticities at separation distances between 0.1 and 2 arcminutes. The cross-component of the correlation function shows no significant deviation from zero as expected, as gravitational lensing only shears galaxies tangentially. For the $\xi_+$ component, the ray-traced (plotted in green) and halo model (plotted in purple) are in relative agreement from seperation scales of 0.1 to 0.2 arcminutes, but there is a clear systematic underprediction of correlation by the halo model on scales larger than 0.2 arcminutes. This may be indicative that not enough halo mass is being used in the lensing prediction, the halo model does not adequately address large-scale mass structures, or that there is significant dark matter mass not correlated with galaxies. These issues are discussed further in Section \ref{issues}.

Using the same catalog of background sources, this statistic was recalculated for the halo model ellipticities for various lightcone radii ranging from $R_L=1$ to $R_L=8$ arcminutes. The result is shown in Figure \ref{fig:gg_corr_series}, where the series of model predicted correlation functions is compared to the ray-traced correlation function. As $R_L$ increases, the mean number of foreground galaxies contained in each lightcone grows quadratically from ${134\pm17}$ galaxies when $R_L=1$ arcminute to ${8570\pm87}$ galaxies when $R_L=8$ arcminutes. Increasing the number of foreground objects, and thus increasing the mass and structure considered for shear and convergence predictions, systematically increased the predicted correlation on all scales. This result encourages the use of lightcones with radii of at least 8 arcminutes, and further emphasizes the need for computational efficiency to allow the timely use of large lightcones.

@@

The catalog of lensed ellipticities was also analyzed using the galaxy-mass correlation function with separation distances from 0.1 to 2 arcminutes and a lightcone radius of 6 arcminutes. The result is shown in Figure \ref{fig:ng_corr}. The lensed ellipticities from the ray-tracing and halo model both show positive correlation on all calculated scales and largely agree on scales larger than 0.3 arcminutes except for a slight overprediction by the halo model near 0.45 arcminutes. However, a new feature is the large overprediction of correlation on separation scales smaller than 0.2 arcminutes. This is likely because background sources this close to massive foreground halos are often in the strong lensing regime which is not currently accounted for by the \texttt{Pangloss} framework.

The galaxy-mass correlation function was similarly re-calculated at increasing lightcone radii from $R=1$ to $R=8$ arcminutes with the result in Figure \ref{fig:ng_corr_series} As with the $\varepsilon$-$\varepsilon$ correlation, structure larger than $\sim$1 arcminute is captured much better by increasingly large lightcone radii as expected. However, there is little to no improvement of the galaxy-mass correlation on scales smaller than 0.3 arcminutes which futher points to the need to model strong lensing.

%%------------------------------------------------------------------------------------------------------------------------------------------
%% Galaxy-mass correlation function
\begin{figure}
    \centering
    \begin{subfigure}{0.475\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/ng_corr.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{CAPTION}
        %The galaxy-mass correlation function of 3,240 sources at separation distances from 0.1 to 2 arcminutes and a lightcone radius of $R=2$ arcminutes. There is no significant correlation of the intrinsic ellipticities (blue) as expected, and both the ray-traced correlation (green) and halo model correlation (purple) are positive on all scales. The green and purple curves are in agreement from 0.2 to 0.7 arcminutes, but have significant discrepancies on scales above and below this range.}
        \label{fig:ng_corr}
    \end{subfigure}
    ~~
    \begin{subfigure}{0.475\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/ng_progression.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{CAPTION}
        \label{fig:ng_corr_series}
    \end{subfigure}
    \caption{}
\end{figure}
%
%%------------------------------------------------------------------------------------------------------------------------------------------

@@

Of intrinsic interest is the relative number of relevant halos needed in each lightcone to accurately predict the halo correlation functions in Figures \ref{fig:gg_corr} and \ref{fig:ng_corr}, as a low percentage would allow much faster lensing predictions. Figures \ref{fig:rel_gg_compare} and \ref{fig:rel_ng_compare} display an example $\varepsilon$-$\varepsilon$ and galaxy-mass correlation function comparrison of the halo model using all halos (purple) and halo model using halos above the relevance limit of ${10^{-5} M_{\odot}/\text{Mpc}^3}$, as well as the percent error of each predicted value between the two implementations. This relevance limit resulted in $??\pm??$ relevant halos per lightcone, and resulted in a normalized root-mean-square error (NRMSE) of ??. Figure \ref{fig:halo_fraction} plots the $\varepsilon$-$\varepsilon$ and galaxy-mass correlation function NRMSE as a function of relevance limits and mean relevant halos. These results suggest that either correlation function prediction can be approximated with an NRMSE below 10\% using around the ?? most relevant halos, or an NRMSE of below 5\% using around the ?? most relevant halos. Calibration curves such as this can allow the best-case correlation function to be predicted within a desired accuracy at only a fraction of the full lensing calculation which will be especially helpful in future analyses with much larger fields of view.

@@

%------------------------------------------------------------------------------------------------------------------------------------------
% NRMSE and Correlation Fraction as function of Mean Relevant Halos
\begin{figure*}
    \centering
    \includegraphics[width=0.925\textwidth]{figs-swe/thesis/halo_fraction_long.png}
    % \includegraphics[width=0.475\textwidth]{figs-swe/thesis/halo_fraction.png}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{CAPTION!}
    \label{fig:halo_fraction}
\end{figure*}

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Convergence Mass Maps} \label{mass_maps}

The result of a particular 9 arcmin$^2$ field is shown in Figure \ref{fig:kappamaps}, where the ray-traced convergence is shown on the left, the halo model convergence is in the center, and the ray-traced convergence subtracted from the halo model convergence is shown on the right. While particular foreground masses are almost always predicted, the currently implemented halo model over-predicts the convergence throughout most of the field; b. The halo model does not appear to accurately account for the voids between foreground masses and over-predicts the mass of most ... This could partially be a result of incorrect halo profiles, but it is apparent that the smooth-component correction is not adequate to ...

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Computational Performance} \label{comp_perform}

%%------------------------------------------------------------------------------------------------------------------------------------------
% Mean lightcone CPU time
\begin{figure}
    \centering
    \includegraphics[width=0.475\textwidth]{figs-swe/thesis/profile_bar.png}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{OLD: The mean CPU runtime per lightcone for various runs with different (and cumulative) code optimizations described in the text. The cumulative changes account for a 59\% decrease in the mean CPU time and a halving of the variation. Currently, the prediction can be made for a single lightcone in $57\pm8$ ms.}
    \label{fig:cpu_plot}
\end{figure}
%%------------------------------------------------------------------------------------------------------------------------------------------

A secondary goal of this research was to investigate how quickly the shear and convergence prediction for a single background source could be made. The mean CPU time required for the lensing prediction of a single lightcone with radius $R_L=6$ arcminutes was ${140\pm20}$ ms for the first run. However, many small improvements to the coding infrastructure of \texttt{Pangloss} were made to speed up the prediction. The mean CPU time for a single lightcone lensing prediction for 6 separate runs after various (and cumulative) changes is plotted in Figure \ref{fig:cpu_plot}, with a final run time of $57\pm8$ ms. Run 1 corresponds to the original \texttt{Pangloss} framework code, run 2 wrote only the needed quantities for this analysis to the catalog \texttt{Astropy}\footnote{http://www.astropy.org/} tables, run 3 changed the redshift grid binning to be done once for all halos in the drilling rather than for every individual halo in lensing, run 4 optimized various lensing calculations with matrices whenever possible, run 5 replaced the computationally expensive lensing calculation functions with a lookup table, and run 6 used only the ${65\pm9}$ most relevant halos from a relevance limit of ${10^{-5}M_\odot/\text{Mpc}^3}$. Cumulatively, these changes accounted for a 59\% decrease in mean CPU time per lightcone and halved the initial variation. While significant progress has been made, the code needs to be further optimized to allow lensing predictions of fields of many square degrees to be done in a few seconds. Section \ref{scaling_up} discusses the details of future computational scale-up plans.

Figure \ref{fig:cpu_scaling} displays how the mean CPU runtime per lightcone scales as a function of lightcone radius $R_L$ with all halos (left) and relevance limit at a constant radius of $R_L=4$ and $R_L=8$ arcminutes (right). Unsurprisingly, the runtime depends strongly on the number of galaxies per lightcone. Both figures show that the lensing calculation requires an overhead of about ${55\pm10}$ ms, but then the calculation speed depends quadratically on the number of halos and logarithmically on the relevance limit. The runtime saved by using only the most relevant halos helps can only counteract the increase in runtime due to larger lightcones for so long before the desired correlation function NRMSE becomes too large, so accurate analysis with radii above $R_L=8$ arcminutes may only be feasible with parallelization.

The distribution of total CPU time in significant parts of the lensing calculation for run 6 are as follows: 24.0\% of total time calculating individual halo convergence contributions where 14.4\% is spent in lensing lookup tables, 65.1\% of CPU time was spent combining all convergence contributions where 62.1\% is spent on the smooth-component correction, and the majority of the remaining time spent on writing the output \texttt{Astropy} columns. However, time spent accross all components of the lensing calculation was dominated by \texttt{Astropy} column use with nearly 55\% of total time spent initializing, copying, or summing columns.

%%------------------------------------------------------------------------------------------------------------------------------------------
%% Mean CPU time scaling
\begin{figure*}
    \centering
    \includegraphics[width=0.8\textwidth]{figs-swe/thesis/cpu_scaling.png}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{CAPTION}
    \label{fig:cpu_scaling}
\end{figure*}
%
%%------------------------------------------------------------------------------------------------------------------------------------------

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
\subsection{Results} \label{inference_results}
